### 2025-03-13 paper/sections/intro.qmd
some work on cutting content

Part of the problem is that the high degrees of freedom provide room for many solutions that are locally optimal with respect to narrow objectives [@wilson2020case].^[We follow the standard ML convention, where "degrees of freedom" refer to the number of parameters estimated from data.] Indeed, recent work on the so called "lottery ticket hypothesis" suggests that modern neural networks can be pruned by up to 90% while preserving their predictive performance [@frankle2018lottery] and generalizability [@morcos2019success]. Similarly, @zhang2021understanding showed that state-of-the-art neural networks are so expressive that they can fit randomly labeled data. Thus, looking at the predictive performance alone, the solutions may seem to provide compelling explanations for the data, when in fact they are based on purely associative, semantically meaningless patterns. This poses two related challenges. Firstly, there is no dependable way to verify if such complex representations correspond to meaningful and plausible explanations. Secondly, even if we could resolve the first challenge, it remains undecided how to ensure that models can *only* learn valuable explanations.

@@

**Specifically, we introduce counterfactual training**: a novel training regime explicitly meant to align learned representations with plausible explanations that comply with user requirements. Our contributions are as follows:

- We present a novel methodological framework that leverages adversarial examples and faithful counterfactual explanations during the training phase to improve the explanatory capacity and robustness of machine learning models (@sec-method).  
- We propose a method to enforce global actionability constraints by preventing models from assigning importance to immutable features, i.e., ones over which decision subjects have no control (@sec-method).  
- Through extensive experiments we demonstrate that counterfactual training promotes explainability while preserving high predictive performance. We run ablation studies and grid searches to understand how the underlying model components and hyperparameters affect outcomes. (@sec-experiments).

Despite some limitations discussed in @sec-discussion, we conclude in @sec-conclusion that counterfactual training provides a useful framework for researchers and practitioners interested in making opaque models more trustworthy. We also believe that this work serves as an opportunity for XAI researchers to re-evaluate the trend of improving XAI tools without improving the underlying models.

>>>>

Part of the problem is that the high degrees of freedom provide room for many solutions that are locally optimal with respect to narrow objectives [@wilson2020case].^[We follow the standard ML convention, where "degrees of freedom" refer to the number of parameters estimated from data.] Indeed, recent work on the so-called "lottery ticket hypothesis" suggests that modern neural networks can be pruned by up to 90% while preserving their predictive performance [@frankle2018lottery] and generalizability [@morcos2019success]. Similarly, @zhang2021understanding showed that state-of-the-art neural networks are so expressive that they can fit randomly labeled data. Thus, looking at the predictive performance alone, the solutions may seem to provide compelling explanations for the data, when in fact they are based on purely associative, semantically meaningless patterns. This poses two related challenges. Firstly, there is no dependable way to verify if such complex representations correspond to meaningful and plausible explanations. Secondly, even if we could resolve the first challenge, it remains undecided how to ensure that models can *only* learn valuable explanations.

@@

**Specifically, we introduce counterfactual training (CT)**: a novel training regime explicitly meant to align learned representations with plausible explanations that comply with user requirements. The remainder of this paper is structured as follows. @sec-lit presents related work, focusing in particular on the link between adversarial examples and counterfactual explanations. Then follow our main contributions:

1. In @sec-method, we introduce our methodological framework and show theoretically that it can be used to enforce global actionability constraints.  
2. Through extensive experiments we demonstrate that CT substantially improves explainability without sacrificing predictive performance (@sec-experiments).

We discuss future research and challenges in @sec-discussion and conclude in @sec-conclusion that CT is a promising new approach towards making opaque models more trustworthy.

### 2025-03-13 paper/sections/method.qmd
some work on cutting content

Counterfactual training (CT) combines ideas from adversarial training, energy-based modelling and counterfactuals explanations with the explicit goal of aligning representations with plausible explanations that comply with user requirements. In the context of CEs, plausibility has broadly been defined as the degree to which counterfactuals comply with the underlying data-generating process [@poyiadzi2020face;@guidotti2022counterfactual;@altmeyer2024faithful]. Plausibility is a necessary but insufficient condition for using CEs to provide algorithmic recourse (AR) to individuals (negatively) affected by opaque models. For AR recommendations to be actionable, they need to not only result in plausible counterfactuals but also be attainable. A plausible CE for a rejected 20-year-old loan applicant, for example, might reveal that their application would have been accepted, if only they were 20 years older. Ignoring all other features, this would comply with the definition of plausibility if 40-year-old individuals were in fact more credit-worthy on average than young adults. But of course this CE does not qualify for providing actionable recourse to the applicant since *age* is not a (directly) mutable feature. CT aims to improve model explainability by aligning models with counterfactuals that meet both desiderata: plausibility and actionability. Formally, we define explainability as follows:

@@

$$
\begin{split}
\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) + \lambda_{\text{div}} \text{div}(\mathbf{x},\mathbf{x}_T^\prime,y;\theta) &+ \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y}) \\ &+ \lambda_{\text{reg}}\text{ridge}(\mathbf{x},\mathbf{x}_T^\prime,y;\theta)
\end{split}
$$ {#eq-obj}

where $\text{yloss}(\cdot)$ is a classification loss that induces discriminative performance (e.g., cross-entropy). The second and third terms in @eq-obj are explained in detail below. For now, they can be sufficiently described as inducing explainability directly and indirectly by penalizing: (1) the contrastive divergence, $\text{div}(\cdot)$, between mature counterfactuals $\mathbf{x}_T^\prime$ and observed samples $x$ and, (2) the adversarial loss, $\text{advloss}(.)$, with respect to nascent counterfactuals $\mathbf{x}_{t\leq T}^\prime$. Finally, $\text{ridge}(\cdot)$ denotes a Ridge penalty ($\ell_2$-norm) that regularizes the magnitude of the energy terms involved in $\text{div}(\cdot)$ [@du2019implicit]. The trade-off between the components can be governed by adjusting the strengths of the penalties $\lambda_{\text{div}}$, $\lambda_{\text{adv}}$ and $\lambda_{\text{reg}}$.

@@

$$
\text{div}(\mathbf{x},\mathbf{x}_T^\prime,y;\theta) = \mathcal{E}_\theta(\mathbf{x},y) - \mathcal{E}_\theta(\mathbf{x}_T^\prime,y)
$$ {#eq-div}

where $\mathcal{E}_\theta(\cdot)$ denotes the energy function. We set $\mathcal{E}_\theta(\mathbf{x},\mathbf{y})=-\mathbf{M}_\theta(\mathbf{x^+})[y^+]$ where $y^+$ denotes the index of the randomly drawn target class, $y^+ \sim p(y)$, and $\mathbf{x^+}$ denotes an observed sample from target domain: $\mathbf{X}^+=\{\mathbf{x}:y=y^+\}$. Conditional on the target class $y^+$, $\mathbf{x}_T^\prime$ denotes a mature counterfactual for a randomly sampled factual from a non-target class generated with a gradient-based CE generator for up to $T$ iterations. Mature counterfactuals are ones that have either reached convergence wrt. the decision threshold $\tau$ or exhausted $T$.

@@

Since such actionability constraints are binding at test time, we should also impose them when generating $\mathbf{x}^\prime$ during each training iteration to inform model representations. Through their effect on $\mathbf{x}^\prime$, both types of constraints influence model outcomes via @eq-div. Here it is crucial that we avoid penalizing implausibility that arises due to mutability constraints. For any mutability-constrained feature $d$ this can be achieved by enforcing $\mathbf{x}[d] - \mathbf{x}^\prime[d]:=0$ whenever perturbing $\mathbf{x}^\prime[d]$ in the direction of $\mathbf{x}[d]$ would violate mutability constraints. Specifically, we set $\mathbf{x}[d] := \mathbf{x}^\prime[d]$ if:

1. Feature $d$ is strictly immutable in practice.
2. We have $\mathbf{x}[d]>\mathbf{x}^\prime[d]$, but feature $d$ can only be decreased in practice.
3. We have $\mathbf{x}[d]<\mathbf{x}^\prime[d]$, but feature $d$ can only be increased in practice.

From a Bayesian perspective, setting $\mathbf{x}[d] := \mathbf{x}^\prime[d]$ can be understood as assuming a point mass prior for $p(\mathbf{x})$ with respect to feature $d$. Intuitively, we think of this simply in terms ignoring implausibility costs with respect to immutable features, which effectively forces the model to instead seek plausibility with respect to the remaining features. This in turn results in lower overall sensitivity to immutable features, which we demonstrate empirically for different classifiers in @sec-experiments. Under certain conditions, this results holds theoretically:^[For the proof, see the supplementary appendix.]

@@

In panel (a), we have trained our model conventionally, and we do not impose mutability constraints at test time. The generated counterfactuals are all valid, but not plausible: they do not comply with the distribution of the factual samples in the target class to the point where they are clearly distinguishable from the ground-truth data. In panel (b), we have trained our model with counterfactual training, once again without any mutability constraints. We observe that the counterfactuals are highly plausible, meeting the first objective of Def. \ref{def-explainability}.

In panel (c), we have used conventional training again, this time imposing the mutability constraint on *age* at test time. Counterfactuals are valid but involve some substantial reductions in *debt* for some individuals (very young applicants). By comparison, counterfactual paths are shorter on average in panel (d), where we have used counterfactual training and protected the immutable feature as described in @sec-constraints. We observe that due to the classifier's lower sensitivity to *age*, recourse recommendations with respect to *debt* are much more homogenous and do not disproportionately punish younger individuals. The counterfactuals are also plausible with respect to the mutable feature. Thus, we consider the model in panel (d) as the most explainable according to Def. \ref{def-explainability}.

>>>>

Counterfactual training combines ideas from adversarial training, energy-based modelling and counterfactuals explanations with the explicit goal of aligning representations with plausible explanations that comply with user requirements. In the context of CEs, plausibility has broadly been defined as the degree to which counterfactuals comply with the underlying data-generating process [@poyiadzi2020face;@guidotti2022counterfactual;@altmeyer2024faithful]. Plausibility is a necessary but insufficient condition for using CEs to provide algorithmic recourse (AR) to individuals (negatively) affected by opaque models. For AR recommendations to be actionable, they need to not only result in plausible counterfactuals but also be attainable. A plausible CE for a rejected 20-year-old loan applicant, for example, might reveal that their application would have been accepted, if only they were 20 years older. Ignoring all other features, this would comply with the definition of plausibility if 40-year-old individuals were in fact more credit-worthy on average than young adults. But of course this CE does not qualify for providing actionable recourse to the applicant since *age* is not a (directly) mutable feature. CT aims to improve model explainability by aligning models with counterfactuals that meet both desiderata: plausibility and actionability. Formally, we define explainability as follows:

@@

$$
\begin{split}
\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) + \lambda_{\text{div}} \text{div}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta) &+ \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y}) \\ &+ \lambda_{\text{reg}}\text{ridge}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta)
\end{split}
$$ {#eq-obj}

where $\text{yloss}(\cdot)$ is a classification loss that induces discriminative performance (e.g., cross-entropy). The second and third terms in @eq-obj are explained in detail below. For now, they can be sufficiently described as inducing explainability directly and indirectly by penalizing: (1) the contrastive divergence, $\text{div}(\cdot)$, between mature counterfactuals $\mathbf{x}_T^\prime$ and observed samples $\mathbf{x}^+\in\mathcal{X}^+=\{\mathbf{x}:y=y^+\}$ in the target class $y^+$, and, (2) the adversarial loss, $\text{advloss}(.)$, with respect to nascent counterfactuals $\mathbf{x}_{t\leq T}^\prime$. Finally, $\text{ridge}(\cdot)$ denotes a Ridge penalty ($\ell_2$-norm) that regularizes the magnitude of the energy terms involved in $\text{div}(\cdot)$ [@du2019implicit]. The trade-off between the components can be governed by adjusting the strengths of the penalties $\lambda_{\text{div}}$, $\lambda_{\text{adv}}$ and $\lambda_{\text{reg}}$.

@@

$$
\text{div}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta) = \mathcal{E}_\theta(\mathbf{x}^+,y) - \mathcal{E}_\theta(\mathbf{x}_T^\prime,y)
$$ {#eq-div}

where $\mathcal{E}_\theta(\cdot)$ denotes the energy function. We set $\mathcal{E}_\theta(\mathbf{x},y)=-\mathbf{M}_\theta(\mathbf{x})[y^+]$ where $y^+$ denotes the index of the randomly drawn target class, $y^+ \sim p(y)$. Conditional on the target class $y^+$, $\mathbf{x}_T^\prime$ denotes a mature counterfactual for a randomly sampled factual from a non-target class generated with a gradient-based CE generator for up to $T$ iterations. Mature counterfactuals are ones that have either reached convergence wrt. the decision threshold $\tau$ or exhausted $T$.

@@

Since such actionability constraints are binding at test time, we should also impose them when generating $\mathbf{x}^\prime$ during each training iteration to inform model representations. Through their effect on $\mathbf{x}^\prime$, both types of constraints influence model outcomes via @eq-div. Here it is crucial that we avoid penalizing implausibility that arises due to mutability constraints. For any mutability-constrained feature $d$ this can be achieved by enforcing $\mathbf{x}^+[d] - \mathbf{x}^\prime[d]:=0$ whenever perturbing $\mathbf{x}^\prime[d]$ in the direction of $\mathbf{x}^+[d]$ would violate mutability constraints. Specifically, we set $\mathbf{x}^+[d] := \mathbf{x}^\prime[d]$ if:

1. Feature $d$ is strictly immutable in practice.
2. We have $\mathbf{x}^+[d]>\mathbf{x}^\prime[d]$, but feature $d$ can only be decreased in practice.
3. We have $\mathbf{x}^+[d]<\mathbf{x}^\prime[d]$, but feature $d$ can only be increased in practice.

From a Bayesian perspective, setting $\mathbf{x}^+[d] := \mathbf{x}^\prime[d]$ can be understood as assuming a point mass prior for $p(\mathbf{x}^+)$ with respect to feature $d$. Intuitively, we think of this simply in terms ignoring implausibility costs with respect to immutable features, which effectively forces the model to instead seek plausibility with respect to the remaining features. This in turn results in lower overall sensitivity to immutable features, which we demonstrate empirically for different classifiers in @sec-experiments. Under certain conditions, this results holds theoretically:^[For the proof, see the supplementary appendix.]

@@

In panel (a), we have trained our model conventionally, and we do not impose mutability constraints at test time. The generated counterfactuals are all valid, but not plausible: they do not comply with the distribution of the factual samples in the target class to the point where they are clearly distinguishable from the ground-truth data. In panel (b), we have trained our model with CT, once again without any mutability constraints. We observe that the counterfactuals are highly plausible, meeting the first objective of Def. \ref{def-explainability}.

In panel (c), we have used conventional training again, this time imposing the mutability constraint on *age* at test time. Counterfactuals are valid but involve some substantial reductions in *debt* for some individuals (very young applicants). By comparison, counterfactual paths are shorter on average in panel (d), where we have used CT and protected the immutable feature as described in @sec-constraints. We observe that due to the classifier's lower sensitivity to *age*, recourse recommendations with respect to *debt* are much more homogenous and do not disproportionately punish younger individuals. The counterfactuals are also plausible with respect to the mutable feature. Thus, we consider the model in panel (d) as the most explainable according to Def. \ref{def-explainability}.

### 2025-03-13 paper/sections/other/grids.qmd
some work on cutting content

Overall, we find that Counterfactual Training (CT) achieves it key objectives consistently across all hyperparameter settings and also broadly across datasets: plausibility is improved by up to ~60 percent (%) for the *Circles* data (e.g. @fig-grid-gen_params-plaus-circles), ~25-30% for the *Moons* data (e.g. @fig-grid-gen_params-plaus-moons) and ~10-20% for the *Linearly Separable* data (e.g. @fig-grid-gen_params-plaus-lin_sep). At the same time, the average costs of faithful counterfactuals are reduced in many cases by around ~20-25% for  *Circles* (e.g. @fig-grid-gen_params-cost-circles) and up to ~50% for *Moons* (e.g. @fig-grid-gen_params-cost-moons). For the *Linearly Separable* data, costs are generally increased although typically by less than 10% (e.g. @fig-grid-gen_params-cost-lin_sep), which reflects a common tradeoff between costs and plausibility [@altmeyer2024faithful].

>>>>

Overall, we find that counterfactual training (CT) achieves it key objectives consistently across all hyperparameter settings and also broadly across datasets: plausibility is improved by up to ~60 percent (%) for the *Circles* data (e.g. @fig-grid-gen_params-plaus-circles), ~25-30% for the *Moons* data (e.g. @fig-grid-gen_params-plaus-moons) and ~10-20% for the *Linearly Separable* data (e.g. @fig-grid-gen_params-plaus-lin_sep). At the same time, the average costs of faithful counterfactuals are reduced in many cases by around ~20-25% for  *Circles* (e.g. @fig-grid-gen_params-cost-circles) and up to ~50% for *Moons* (e.g. @fig-grid-gen_params-cost-moons). For the *Linearly Separable* data, costs are generally increased although typically by less than 10% (e.g. @fig-grid-gen_params-cost-lin_sep), which reflects a common tradeoff between costs and plausibility [@altmeyer2024faithful].

### 2025-03-13 paper/sections/related.qmd
some work on cutting content

Once again it helps to look at these findings through the lens of representation learning with high degrees of freedom. Deep ensembles are approximate Bayesian model averages, which are most called for when models are underspecified by the available data [@wilson2020case]. Averaging across solutions mitigates the aforementioned risk of relying on a single locally optimal representations that corresponds to semantically meaningless explanations for the data. Previous work by @schut2021generating similarly found that generating plausible ("interpretable") counterfactual explanations is almost trivial for deep ensembles that have also undergone adversarial training. The case for JEMs is even clearer: they involve a hybrid objective that induces both high predictive performance and generative capacity [@grathwohl2020your]. This is closely related to the idea of aligning models with plausible explanations and has inspired our proposed counterfactual training objective, as we explain in @sec-method.

>>>>

Once again it helps to look at these findings through the lens of representation learning with high degrees of freedom. Deep ensembles are approximate Bayesian model averages, which are most called for when models are underspecified by the available data [@wilson2020case]. Averaging across solutions mitigates the aforementioned risk of relying on a single locally optimal representations that corresponds to semantically meaningless explanations for the data. Previous work by @schut2021generating similarly found that generating plausible ("interpretable") counterfactual explanations is almost trivial for deep ensembles that have also undergone adversarial training. The case for JEMs is even clearer: they involve a hybrid objective that induces both high predictive performance and generative capacity [@grathwohl2020your]. This is closely related to the idea of aligning models with plausible explanations and has inspired our proposed CT objective, as we explain in @sec-method.
